Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2024-04-18 doc/manuscript/translocation_bioarxiv_SI.tex
Update PNAS template
Add new figure, text edits, and latex files

\documentclass[9pt,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

@@

We estimated survival and recruitment for each site using open
population CMR models based on \citep{joseph2018}. For each individual
\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
the individual was not detected, and \(o_{i, j}=2\) if the individual
was detected. Capture histories of \(M\) individuals are modeled,
although only \(N_s\) individuals were captured. This parameter expanded
data augmentation allows us to capture the possibility that undetected
individuals may have recruited into the adult population
\citep{royle2012}. Here, \(M\) was chosen to be three times the number
of observed individuals (\(3N_s\)) to be considerably greater than our
prior guess of \(N_s\).

We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
period \(t = 1,..., n_t\). The four states that we consider are:
\(u_{i, t} = 1\) for individuals that have not recruited,
\(u_{i, t} =2\) for live adults, and \(u_{i, t} = 3\) for dead adults.
Each survey \(j=1, ..., n_j\) occurs in one of the \(n_t\) primary
periods, and we denote the primary period in which survey \(j\) takes
place as \(t_j\). Each primary period \(t\) occurs within one year, but
within a year there can be multiple primary periods. We set the year
containing the first primary period to \(y_{t = 1} = 1\), and generally
\(y_t\) represents the year containing primary period \(t\). Years
increment by one until the final year of the mark recapture efforts,
which we denote \(n_y\): \(y \in \{1, 2, ..., n_y\}\). We assume that
within a primary period, the state of each individual does not change
(i.e., individuals do not recruit into the adult population, gain or
lose Bd infection, or die). This assumption is justified by the short
time intervals between surveys within primary periods, in cases where
primary periods contain multiple secondary periods.

@@

\[
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 1 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
\] where \(\lambda_t\) is the probability of recruiting in time \(t\)
and \(\phi_t\) is the probability of survival in time \(t\).

@@

\[
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 1 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
\] where \(I_{t, i}\) is a known indicator function for whether
individual \(i\) was introduced in primary period \(t\).

@@

To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
predicted true states for naturally-recruited individuals (1=not
recruited, 2=alive, 3=dead, as described in \textbf{Supporting
Information - Frog population recovery - CMR model structure}), then
calculated the posterior probability of individuals surviving between
consecutive primary periods, conditional on being alive in the first
primary period (e.g., given a value of 2 (alive) in the first primary
period, how often was the value still 2 (alive) in the next primary
period compared to 3 (dead)?) . This yielded posterior distributions for
survival probabilities between primary periods. However, because the
time interval between primary periods differed, the survival
probabilities between different consecutive primary periods were not
directly comparable. To address this, we converted the survival
probabilities between each consecutive primary period to per day death
rates, propagating the uncertainty from the posterior distributions. We
then took a weighted average of these death rates, weighted by the time
interval between primary periods, to get the average per day death rate
over the entire CMR survey. We converted this per day death rate \(d\)
back to a yearly survival probability using
\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
\(\sigma_{A_T}\) such that our estimates of average yearly survival
probability were comparable between \(\sigma_{A_R}\) and
\(\sigma_{A_T}\).

@@

Second, we performed a local elasticity analysis on \(\lambda\) with
respect to parameters \(\sigma_{J_1}\), \(\sigma_{J_2}\),
\(\sigma_{A_R}\), and \(F\) to determine how small changes in these
parameters could influence the long-run deterministic growth rate of
populations (Figure~\ref{fig-viability-supp}).

@@

\hypertarget{frog-evolution-in-response-to-bd-2}{%
\section{Frog evolution in response to
Bd}\label{frog-evolution-in-response-to-bd-2}}

@@

\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
the 12 recipient sites, Bd loads for the period immediately prior to
translocation versus during the 1-year period after translocation. Bd
loads are expressed as the number of ITS1 copies per skin swab, as
estimated by qPCR of the Bd ITS1 region. Box plots show medians, first
and third quartiles, largest and smallest values within 1.5x
interquartile range, and values outside the 1.5x interquartile range.
Loads indicative of severe disease are \textgreater{} 5.8 ITS copies (on
a log\textsubscript{10} scale). Samples sizes are provided immediately
above the x-axis.}

@@

\caption{\label{fig-viability-supp}Sensitivity analysis of the
stage-structured MYL frog model. Elasticity of \(\lambda\) with changes
in four parameters: \(\sigma_{A_R}\) (yearly survival probability of
naturally recruited adults), \emph{F (}number of eggs produced by a
female frog in a year that successfully hatch), \(\sigma_{J_1}\) (yearly
probability of survival of year-1 juveniles that also affects
recruitment), and \(\sigma_{J_2}\) (yearly probability of survival and
recruitment of year-2 juveniles). Elasticity is calculated at the
default parameter values for each population and \(\sigma_{J_1}\) =
0.09.}

@@

\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
windows identified by the overlapping splined windows for FST and
\(\pi\), boxplots showing average \(\pi\) for naive and recovering
populations. The corresponding gene annotations for each window are as
follows: (1) C6, C7, (2) DDX10, ZBTB24, (3) SULTR-like, TRAF3IP2, VGLL3,
EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1, TCF19, HSP70, LSM2, VARS1, (5)
GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1, (7) PPP1R12A, TSPAN4, PAWR,
MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6, BSPRY, MPV17, and (9) CAD,
ATRAID, GPN1.}

@@

\hypertarget{tables}{%
\subsubsection{Tables}\label{tables}}

>>>>

\documentclass[9pt,twoside,lineno]{pnas-new-SI}
% Use the lineno option to display guide line numbers if required.

@@

We estimated survival and recruitment for each site using open
population CMR models based on \citep{joseph2018}. For each individual
in a superpopulation of real and psuedo-individuals, \(i=1, ..., M\) on
each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
histories of \(M\) individuals are modeled, although only \(N_s\)
individuals were captured. This parameter expanded data augmentation
allows us to capture the possibility that undetected individuals may
have recruited into the adult population \citep{royle2012}. Here, \(M\)
was chosen to be three times the number of observed individuals
(\(3N_s\)) to be considerably greater than our prior guess of \(N_s\).

We denote the true state of individual \(i\) as \(u_{i, t}\) for primary
period \(t = 1,..., n_t\). The states that we consider are:
\(u_{i, t} = 1\) for individuals that have not recruited,
\(u_{i, t} =2\) for live adults, and \(u_{i, t} = 3\) for dead adults.
Each survey \(j=1, ..., n_j\) occurs in one of the \(n_t\) primary
periods, and we denote the primary period in which survey \(j\) takes
place as \(t_j\). Multiple primary periods can occur during a summer
active season (typically July-August, but in some years as early as May
and as late as September). We set the year containing the first primary
period to \(y_{t = 1} = 1\), and generally \(y_t\) represents the year
containing primary period \(t\). Years increment by one until the final
year of the mark recapture efforts, which we denote \(n_y\):
\(y \in \{1, 2, ..., n_y\}\). We assume that within a primary period,
the state of each individual does not change (i.e., individuals do not
recruit into the adult population, gain or lose Bd infection, or die).
This assumption is justified by the short time intervals between surveys
within primary periods, in cases where primary periods contain multiple
secondary periods.

@@

\[
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 0 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
\] where \(\lambda_t\) is the probability of recruiting in time \(t\)
and \(\phi_t\) is the probability of survival in time \(t\).

@@

\[
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 0 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
\] where \(I_{t, i}\) is a known indicator function for whether
individual \(i\) was introduced in primary period \(t\).

@@

To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
predicted true states for naturally-recruited individuals (1=not
recruited, 2=alive, 3=dead, as described in \textbf{Supporting
Information - Frog population recovery - CMR model structure}), then
calculated the posterior probability of individuals surviving between
consecutive primary periods, conditional on being alive in the first
primary period (e.g., given a value of 2 (alive) in the first primary
period, how often was the value still 2 (alive) in the next primary
period compared to 3 (dead)) . This yielded posterior distributions for
survival probabilities between primary periods. However, because the
time interval between primary periods differed, the survival
probabilities between different consecutive primary periods were not
directly comparable. To address this, we converted the survival
probabilities between each consecutive primary period to per day death
rates, propagating the uncertainty from the posterior distributions. We
then took a weighted average of these death rates, weighted by the time
interval between primary periods, to get the average per day death rate
over the entire CMR survey. We converted this per day death rate \(d\)
back to a yearly survival probability using
\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
\(\sigma_{A_T}\) such that our estimates of average yearly survival
probability were comparable between \(\sigma_{A_R}\) and
\(\sigma_{A_T}\).

@@

Second, we performed a local elasticity and sensitivity analysis on
\(\lambda\) with respect to parameters \(\sigma_{L_1}\),
\(\sigma_{L_2}\), \(\sigma_{L_3}\), \(\sigma_{J_1}\), \(\sigma_{J_2}\),
\(\sigma_{A_R}\), \(F\), \(p_{L_2}\), \(p_{J_1}\) to determine how small
changes in these parameters could influence the long-run deterministic
growth rate of populations (Figure~\ref{fig-viability-supp}). Note
that \(\lambda\) is equally or more sensitive to changes in adult
survival (\(\sigma_{A_R}\)) and survival of year-1 juveniles
(\(\sigma_{J_1}\)) than it is to other model parameters, providing
additional justification for focusing on variation in \(\sigma_{A_R}\)
and \(\sigma_{J_1}\) in our viability analyses.

@@

\hypertarget{frog-genome-evolution-2}{%
\section{Frog genome evolution}\label{frog-genome-evolution-2}}

@@

\caption{\label{fig-bdload-beforeafter}For frogs translocated to each of
the 12 recipient sites, Bd loads for the period immediately prior to
translocation versus during the 1-year period after translocation. Bd
loads are expressed as the number of ITS1 copies per skin swab, as
estimated by qPCR of the Bd ITS1 region. Box plots show medians, first
and third quartiles, largest and smallest values within 1.5x
interquartile range, and values outside the 1.5x interquartile range.
Samples sizes are provided immediately above the x-axis. The red dashed
horizontal line indicates the approximate load above which severe
disease is typically observed \citep{vredenburg2010}. This threshold
(5.8 ITS1 copies - on a log\textsubscript{10} scale) is commonly
exceeded during Bd epizootics, but is exceeded only rarely in recovering
populations \citep{briggs2010}.}

@@

\caption{\label{fig-viability-supp}Sensitivity analysis of the
stage-structured MYL frog model. A. Elasticity of \(\lambda\) with
changes in nine parameters: \(\sigma_{A_R}\) (yearly survival
probability of naturally recruited adults), \(F\) (number of eggs
produced by a female frog in a year that successfully hatch),
\(\sigma_{J_1}\) (yearly probability of survival of year-1 juveniles
that also affects recruitment), \(\sigma_{J_2}\) (yearly probability of
survival and recruitment of year-2 juveniles), \(p_{L_2}\) (probability
that year-2 tadpoles metamorph into juveniles), \(p_{J_1}\) (probability
that year-1 juveniles mature to adults), \(\sigma_{L_1}\) (survival
probability of year-1 tadpoles), \(\sigma_{L_2}\) (survival probability
of year-2 tadpoles), and \(\sigma_{L_3}\) (survival probability of
year-3 tadpoles). B. Sensitivity of \(\lambda\) with changes in the same
nine parameters. Note that \(F\) has different units than the other
parameters, which is why its sensitivity is low. Sensitivities of all
other parameters are comparable. We did not include \(p_{L_1}\) as this
parameter is generally 1 across the populations in this study (Table S1)
-- tadpoles spend at least a year as tadpoles before metamorphosing.
Elasticity and sensitivity is calculated at the default parameter values
for each population and \(\sigma_{J_1}\) = 0.09.}

@@

\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
windows identified by the overlapping splined windows for
\emph{F\textsubscript{ST}} and \(\pi\), boxplots showing average \(\pi\)
for naive and recovering populations. The corresponding gene annotations
for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/ave_pi_per_site_yose.png}

\caption{For each study population located in Yosemite National Park
(populations 1-4 shown in Site Map), average \(\pi\) for the 9 outlier
windows (panels A-I) identified in the splined window analysis.
Populations within Yosemite are the focus because they are the most
relevant to the translocated populations described in the paper: donor
populations 1 and 4 were the source of frogs for 20 of the 24
translocations and the donor population for the remaining four
translocations is located adjacent to populations 3 and 4. Bars are
color-coded by population category (naive, recovering), and gene
annotations within each window are listed above each panel. Gene regions
described in the paper are in panel A (Complement C6,C7), panel E
(GVINP1), and panel F (extended MHC region that includes two SNPs
identified by the GEMMA analysis in genes TCF19 and VARS1, indicated by
an *). In the Site Map, the four populations included in panels A-I are
color-coded by population category, and dashed lines delineate the
genetic clusters as defined in Figure~\ref{fig-selectionresults}.}

### 2024-04-18 doc/manuscript/translocation_pnas.tex
Add new figure, text edits, and latex files

\begin{abstract}
Vast alteration of the biosphere by humans is causing a sixth mass
extinction, driven in part by an increase in emerging infectious
diseases. The emergence of the lethal fungal pathogen
(\emph{Batrachochytrium dendrobatidis}; ``Bd'') has devastated global
amphibian biodiversity, with hundreds of species experiencing declines
or extinctions. With no broadly applicable methods available to reverse
these impacts in the wild, the future of many amphibians appears grim.
The once-common mountain yellow-legged (MYL) frog is emblematic of
amphibians threatened by Bd. Although most MYL frog populations are
extirpated following disease outbreaks, some persist and eventually
recover. Frogs in these recovering populations have increased resistance
against Bd infection, consistent with evolution of resistant genotypes
and/or acquired immunity. We conducted a 15-year landscape-scale
reintroduction study and show that frogs collected from recovering
populations and reintroduced to vacant habitats can reestablish
populations despite the presence of Bd. In addition, results from
viability modeling suggest that many reintroduced populations have a low
probability of extinction over 50 years. To better understand the role
of evolution in frog resistance, we compared the genomes of MYL frogs
from Bd-naive and recovering populations. We found substantial
differences between these categories, including changes in immune
function loci that may confer increased resistance, consistent with
evolutionary changes in response to Bd exposure. These results provide a
rare example of how reintroduction of resistant individuals can allow
the landscape-scale recovery of disease-impacted species. This example
has broad implications for the many taxa worldwide that are threatened
with extinction by novel pathogens.
\end{abstract}

@@

\caption{\label{fig-allelefrequencies}Evidence for selection on
individual variants in recovering MYL frog populations at the landscape
scale. For each of the 9 naive and recovering MYL frog populations
(indicated by numbered points), adjacent pie charts show allele
frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3.
Charts are superimposed on a map of the Sierra Nevada study area, with
Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
shown in dark gray, and the range boundary for MYL frogs shown in light
gray. The inset map locates the study area and range boundary in
California. Sites 1 and 4 (site id = 72996 and 70567, respectively) were
also used as sources of frogs in the frog population recovery study.}

@@

\caption{\label{fig-spline-manhattan}Evidence for selection on genomic
regions in recovering MYL frog populations. Manhattan plot of the
results from the splined window analysis showing outlier regions for the
difference in (A) nucleotide diversity \(\pi_{diff}\) and (B)
\emph{F\textsubscript{ST}}. In (A), outlier regions are shown above the
upper red dashed line and below the lower red dashed line. In (B),
outlier regions are shown above the single dashed red line. Outlier
regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
shown in blue and outlier regions for both \(\pi_{diff}\) and
\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
(A) showing two adjacent outlier regions for \(\pi_{diff}\) 12.9Mb
upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
line). (D) Magnified Contig8 from from (B) showing the
\emph{F\textsubscript{ST}} outlier region that includes the outlier SNPs
TCF19 and VARS. This region of the genome contains 8 annotated genes
known to occur in the extended MHC Class I and III regions.}

>>>>

\begin{abstract}
Vast alteration of the biosphere by humans is causing a sixth mass
extinction, driven in part by an increase in infectious diseases. The
emergence of the lethal fungal pathogen (\emph{Batrachochytrium
dendrobatidis}; ``Bd'') has devastated global amphibian biodiversity,
with hundreds of species experiencing declines or extinctions. Lacking
any broadly applicable methods to reverse these impacts, the future of
many amphibians appears grim. The once-common mountain yellow-legged
(MYL) frog is emblematic of amphibians threatened by Bd. Although most
MYL frog populations are extirpated following disease outbreaks, some
persist and eventually recover. Frogs in these recovering populations
have increased resistance against Bd infection, consistent with
evolution of resistant genotypes. We conducted a 15-year landscape-scale
reintroduction study and show that frogs collected from recovering
populations and reintroduced to vacant habitats can reestablish
populations despite the presence of Bd. In addition, the likelihood of
establishment is influenced by site, population, and frog attributes.
Results from viability modeling suggest that many reintroduced
populations have a low probability of extinction over 50 years. To
better understand the role of evolution in frog resistance, we compared
the genomes of MYL frogs from Bd-naive and recovering populations. We
found numerous differences between these categories, including in immune
function loci associated with disease resistance, consistent with
evolutionary changes in response to Bd exposure. These results provide a
rare example of how reintroduction of resistant individuals can allow
the landscape-scale recovery of disease-impacted species, and have broad
implications for the many taxa worldwide that are threatened with
extinction by novel pathogens.
\end{abstract}

@@

\caption{\label{fig-allelefrequencies}Genomic patterns consistent with
selection on individual variants in recovering MYL frog populations at
the landscape scale (based on a linear mixed model analysis). For each
of the 9 naive and recovering MYL frog populations (indicated by
numbered points), adjacent pie charts show allele frequencies for the 11
outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS,
MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map
of the Sierra Nevada study area, with Yosemite, Kings Canyon, and
Sequoia National Parks (from north to south) shown in dark gray, and the
range boundary for MYL frogs shown in light gray. The inset map locates
the study area and range boundary in California. Sites 1 and 4 (site id
= 72996 and 70567, respectively) were also used as sources of frogs in
the frog population recovery study.}

@@

\caption{\label{fig-spline-manhattan}Patterns consistent with selection
on genomic regions in recovering MYL frog populations based on a splined
window analysis. Note that several key candidate genes (RIN3, TCF19,
VARS) are outliers in both the individual variant analysis
Figure~\ref{fig-allelefrequencies} and splined window analysis (this
figure). Manhattan plots show outlier regions for the difference in (A)
nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
In (A), outlier regions are shown above the upper red dashed line and
below the lower red dashed line. In (B), outlier regions are shown above
the single dashed red line. Outlier regions for either \(\pi_{diff}\) or
\emph{F\textsubscript{ST}} are shown in blue and outlier regions for
both \(\pi_{diff}\) and \emph{F\textsubscript{ST}} are shown in red. (C)
Magnified Contig19 from (A) showing two adjacent outlier regions for
\(\pi_{diff}\) 12.9Mb upstream of the RIN3 outlier SNP (indicated with a
dashed vertical blue line). (D) Magnified Contig8 from from (B) showing
the \emph{F\textsubscript{ST}} outlier region that includes the outlier
SNPs TCF19 and VARS. This region of the genome contains 8 annotated
genes known to occur in the extended MHC Class I and III regions.}
